Next Trippy Structures\thanks{Supported by organization x.}

### 2019-01-17 paper.tex
Minor corrections

As \ctr is a complete representation of the collection of trips, we are able to extract any trip using the structures described. \marginpar{The idea here is that explaining how to extract will help understand the more complex FromXtoY query later} The algorithm to extract a trip is shown in the Algorithm~\ref{alg:extract}, where \FuncSty{Rank} and \FuncSty{Select} operate over the bitvector $D$ from (i), \FuncSty{WML}{(s$_a$)} is the WM corresponding to the stop \DataSty{s$_a$} in the optimized version of (ii) and \FuncSty{TrackDown} returns the leaf index of a WM given a root index. In a practical implementation, it is not needed to access (ii) and (iii) for the line identifier and jcode of the last stop of the trip, as they will always match the ones from the previous ones.

@@

A more complex query that we can solve with \ctr is ``number of trips that started from a stop \DataSty{s$_a$} and ended at a stop \DataSty{s$_z$}'', which can be further restricted to specific starting and ending lines and a time window. The pseudocode for the full version of such query can be found at the Algorithm~\ref{alg:xy}. This algorithm relies heavily on the abstract function \FuncSty{GetRange}, which when applied to a CSA, delimits the range of entries of $\Psi$ that match a queried string pattern, while on a WM it reports the range in the last level of a WM corresponding to the entries of a queried symbol within a range in the encoded sequence (equivalent to two \FuncSty{TrackDown} operations on the first and last occurrence of the queried symbol, but much more time efficient)\marginpar{Briefly describe WM in previous works and refer to that section for complexity of these queries}.

@@

    \item In \textbf{line 7} we query (ii) in the $\$$ section, within the range previously obtained, for the queried starting line \DataSty{l$_a$}, obtaining the range of its occurences in the last level, delimited by \DataSty{a} and \DataSty{z}. Note that if the \ctr was constructed without the optimization that separates (ii) in sections, this line would be exactly the same, save for the query being on a \DataSty{WML} that would encode the whole sequence of lines instead of just the \DataSty{WML}{(0)} for $\$$.

@@

    \item In \textbf{lines 10-12} we use the \FuncSty{TrackUp} operation, which is the inverse of \FuncSty{TrackDown}: it returns the root index given a leaf index of a WM. In this case, as the \FuncSty{a} and \FuncSty{z} we obtained in the previous step are also indexes in the last level of the lines WM, we use \FuncSty{TrackUp} to translate that range to indexes to the root of the (ii) and therefore to the entries of (i) as well. Note that this is a range inside the $\$$ section for trips with the same origin stop, final stop and starting line that includes all the trips for jcodes that span from \DataSty{j$_a$} to \DataSty{j$_z$}. The properties of (i) also ensure that this range maintains the same size after translation and it can also be directly translated to a range in the \DataSty{s$_z$}.

@@

    \item In \textbf{line 14} the range \DataSty{a'}$..$\DataSty{z'} is trivially translated to the section for \DataSty{s$_z$} where we query (ii) to obtain the subrange for the line \DataSty{l$_z$}. Remember that \DataSty{WML}{(s$_z$)} representes only the lines for \DataSty{s$_z$}, therefore both the translated indexes and the resulting subrange indexes are relative and must be adjusted by \DataSty{offset}. This would have not been necessary if the optimization was not implemented, and absolute indexes would have been used.

    \item In \textbf{line 15} we obtain the jcode range analagously to line 8, but this time for \DataSty{l$_z$} and \DataSty{s$_z$}.

    \item In \textbf{line 16} we return the number of entries of (iii) between \DataSty{j$_a$} and \DataSty{j$_z$} within our final subrange. The function \FuncSty{GetCount} does not report any range boundaries as \FuncSty{GetRange} does, but simply returns the number of occurences. Of course, the time complexity of this operation is still proportional to the height of the WM even though the matching entries can be scattered (not delimiting a contiguos range).
\end{itemize}

It is trivial to reuse the Algorithm~\ref{alg:xy} to answer queries with less constrains. For example, if we were only interested in restricting the starting line, we could return \DataSty{z}-\DataSty{a} after line 7. Or if we only wanted to restrict the ending line and time, we can do it by skipping the lines 4-12, and using directly \FuncSty{GetRange}{(\FuncSty{WML}{(s$_z$)},left-offset,right-offset,l$_z$)} in line 14. The complexity would increase if we restricted by a time window but not by lines, as would need to iterate through all possible lines for \DataSty{s$_a$} and \DataSty{s$_z$} to obtain the jcodes for each line and perform these operations on (ii) and (iii). Fortunately, the number of lines that a single stop can belong to tends to be rather small in practice, thus with a careful implementation that avoids repeating computation the performance of such query can remain competitive, as will be shown later in Section~\ref{sec:exp}. Finally, to obtain only the trips that started on a given stop we would simply need to set \DataSty{pattern} to $\{0,s_a\}$ in line 2, or alternatively to $\{s_z,0\}$ for the final stop, and skipping the operations on the sections of \DataSty{s$_z$} or $\$$, respectively.

Of course, \ctr can also be used to efficiently obtain other interesting information about trips, such as the top k most boarded stops, as shown in \cite{brisaboa2018compact}, with the additional possibility of differentiating stops that are only used to switch lines in \ctr. However, there is no known way of using this structure to obtain other kinds of information efficiently (e.g. the number of passengers in a journey between two stops).

>>>>

As \ctr is a complete representation of the collection of trips, we are able to extract any trip using the structures described. \marginpar{The idea here is that explaining how to extract will help understand the more complex FromXtoY query later} The algorithm to extract a trip is shown in the Algorithm~\ref{alg:extract}, where \FuncSty{Rank} and \FuncSty{Select} operate over the bitvector $D$ from (i), \FuncSty{WML}{(s$_a$)} is the WM corresponding to the stop \DataSty{s$_a$} in the optimized version of (ii) and \FuncSty{TrackDown} returns the leaf index of a WM given a root index. In a practical implementation, it is not needed to access (ii) and (iii) for the line identifier and jcode of the last stop of the trip, as they will always match the previous ones.

@@

A more complex query that we can solve with \ctr is ``number of trips that started from a stop \DataSty{s$_a$} and ended at a stop \DataSty{s$_z$}'', which can be further restricted to specific starting and ending lines and a time window. The pseudocode for the full version of such query can be found at the Algorithm~\ref{alg:xy}. This algorithm relies heavily on the abstract function \FuncSty{GetRange}, which when applied to a CSA, delimits the range of entries of $\Psi$ that match a queried string pattern, while on a WM it reports the range in the last level of a WM corresponding to the entries of a queried symbol within a range in the encoded sequence (equivalent to two \FuncSty{TrackDown} operations on the first and last occurrence of the queried symbol, but more time efficient)\marginpar{Briefly describe WM in previous works and refer to that section for complexity of these queries}.

@@

    \item In \textbf{line 7} we query (ii) in the $\$$ section, within the range previously obtained, for the queried starting line \DataSty{l$_a$}, obtaining the range of its occurrences in the last level, delimited by \DataSty{a} and \DataSty{z}. Note that if the \ctr was constructed without the optimization that separates (ii) in sections, this line would be exactly the same, save for the query being on a \DataSty{WML} that would encode the whole sequence of lines instead of just the \DataSty{WML}{(0)} for $\$$.

@@

    \item In \textbf{lines 10-12} we use the \FuncSty{TrackUp} operation, which is the inverse of \FuncSty{TrackDown}: it returns the root index given a leaf index of a WM. In this case, as the \FuncSty{a} and \FuncSty{z} we obtained in the previous step are also indexes in the last level of the lines WM, we use \FuncSty{TrackUp} to translate that range of indexes to the root of (ii) and therefore to the entries of (i) as well. Note that this is a range inside the $\$$ section for trips with the same origin stop, final stop and starting line that includes all the trips for jcodes that span from \DataSty{j$_a$} to \DataSty{j$_z$}. The properties of (i) also ensure that this range maintains the same size after translation and it can also be directly translated to a range in the \DataSty{s$_z$}.

@@

    \item In \textbf{line 14} the range \DataSty{a'}$..$\DataSty{z'} is trivially translated to the section for \DataSty{s$_z$} where we query (ii) to obtain the subrange for the line \DataSty{l$_z$}. Remember that \DataSty{WML}{(s$_z$)} represents only the lines for \DataSty{s$_z$}, therefore both the translated indexes and the resulting subrange indexes are relative and must be adjusted by \DataSty{offset}. This would have not been necessary if the optimization was not implemented, and absolute indexes would have been used.

    \item In \textbf{line 15} we obtain the jcode range analogously to line 8, but this time for \DataSty{l$_z$} and \DataSty{s$_z$}.

    \item In \textbf{line 16} we return the number of entries of (iii) between \DataSty{j$_a$} and \DataSty{j$_z$} within our final subrange. The function \FuncSty{GetCount} does not report any range boundaries as \FuncSty{GetRange} does, but simply returns the number of occurrences. Of course, the time complexity of this operation is still proportional to the height of the WM even though the matching entries can be scattered (not delimiting a contiguous range).
\end{itemize}

It is trivial to reuse the Algorithm~\ref{alg:xy} to answer queries with less constrains. For example, if we were only interested in restricting the starting line, we could return \DataSty{z}-\DataSty{a} after line 7. Or if we only wanted to restrict the ending line and time, we can do it by skipping the lines 4-12, and using directly \FuncSty{GetRange}{(\FuncSty{WML}{(s$_z$)},left-offset,right-offset,l$_z$)} in line 14. The complexity would increase if we restricted by a time window but not by lines, as would need to iterate through all possible lines for \DataSty{s$_a$} and \DataSty{s$_z$} to obtain the jcodes for each line and perform these operations on (ii) and (iii). Fortunately, the number of lines that a single stop can belong to tends to be rather small in practice, thus with a careful implementation that avoids repeating computation the performance of such query scales well, as will be shown later in Section~\ref{sec:exp}. Finally, to obtain only the trips that started on a given stop we would simply need to set \DataSty{pattern} to $\{0,s_a\}$ in line 2, or alternatively to $\{s_z,0\}$ for the final stop, and skipping the operations on the sections of \DataSty{s$_z$} or $\$$, respectively.

Of course, \ctr can also be used to efficiently obtain other interesting information about trips, such as the top k most boarded stops, as shown in \cite{brisaboa2018compact}, with the additional possibility of differentiating stops that are only used to switch lines in \ctr. However, there is no known way of using this representation to obtain other kinds of information efficiently (e.g. the number of passengers in a journey between two stops).
